Unscrambling arrested development: temporal ordering and registration of cross-sectional imaging data

### 2014-04-10 image_analysis_paper/pnas_draft.tex
added more edits to paper

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble the image features in Figure~\ref{subfig:images_unordered}.
%
The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
%
We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_n$, the group of integers modulo $n$).
%
Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

@@

We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
%
Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
which make each pair have the maximum possible overlap (see {\it Materials and Methods}).
%%%
%%% so I do not like the figure with the vectors - it is correct, but people have to take the
%%% function, discretize and make it a vector in their heads, and may be "one step too much"
%%%
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
%
From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
%
Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
The problem of finding these alignments is formulated as an optimization problem, and then relaxed to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
provides the rotation angles assigned to each data point (see {\it Materials and Methods}).
%
%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.
%
\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{(a) Set of vectors, each in a different orientation. The pariwise alignment angles are indicated. (b) The vectors from (a), each rotated so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in (a). (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges. (d) The data in (c), colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$. The color intensity is monotonic with the arclength, thus parameterizing the curve.  }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:synch1}{\ref{fig:schematics}a}
\customlabel{subfig:synch2}{\ref{fig:schematics}b}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}c}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}d}
\end{figure}
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
It is reassuring that the peaks are aligned,
even though the algorithm uses no specific information about the existence or location of such peaks.
%
%We note once more that, since angular synchronization only uses pairwise rotation information,
%the optimal rotations are recovered up to a global rotation.

@@

After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
%
%Our main assumption is that most of the variation in the data correlates with time.
%
Our main assumptions are that the high-dimensional data (images have one dimension per color pixel intensity, and concentration profiles have one dimension per discretization point) lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time.
%
Figure~\ref{subfig:dmaps_edges} shows an schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
%
We assume that the data points are ordered from youngest to oldest along the curve.
%
Because our data are very high dimensional, we cannot extract this curve by eye.
%
We instead use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to automatically extract this one-dimensional curve.
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading diffusion map coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect, and ordering the data by the leading diffusion map coordinate will effectively order data in time.

To uncover the diffusion maps parameterization, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that pairs of points which are close have dark/strong edges connecting them.
%
Visually, the dark edges delineate the one-dimensional curve, whereas the light edges are less informative about the low-dimensional structure.
%
%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
%
Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
%
%This procedure has been named diffusion maps \cite{coifman2005geometric}.
%
%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we use diffusion maps to order them in time.
%
We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion maps coordinate, $\phi_2$.
%
As expected, the intensity of the peak grows as a function of $\phi_2$ (since we constructed these data
to have a peak growing in time).
%
%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
%
%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.

@@

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the alignments {\em and} the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
%
The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion maps coordinate.

@@

The same methodologies (with small modifications) can be applied directly to our two-dimensional images.
%
Instead of one-dimensional profiles (discretized to $100$-dimensional concentration vectors), we now work with
two color, two-dimensional images (discretized into $100 \times 100$ pixels/$20000$-dimensional vectors), and we register the images with respect to translations and rotations (the symmetry group is $ISO(2)$).
%
To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\em three-dimensional rotations} (symmetry group $SO(3)$; see {\it Materials and Methods}, {\it SI Appendix}).
%
We use vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
the results are shown in Figure~\ref{fig:images_ordered}.
%
Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of the leading vector diffusion map coordinate.

In these experiments, we are fortunate enough to be able to validate our  automatic registration and ordering using prior physical knowledge about the system.
%
For registration, it is known that the peak in the distribution of Dorsal (shown in green) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
%%%
%%% maybe you can also quantify this guy
%%%
%
%Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position.
%
Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
%
The two methods for registration and ordering can be visually compared through the sequences of one-dimensional profiles extracted from the circumference of the embryos (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
%
The general trends are consistent:
both green  and red peaks are aligned, and the red peaks grow as a function of time/leading vector diffusion map coordinate.
%
However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows an arguably better clustering of the ``outer" peaks in the top corners of the image.
%
%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
%
These peaks are known to arise later in development, and so it is reassuring that they appear at the end of the (recovered) trajectory.
%
%Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
%
Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9158, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig5}
\caption{The images in Figure~\ref{subfig:images_unordered}, aligned and temporally ordered using vector diffusion maps. The images are sorted according to the value of the first non-trivial vector diffusion maps component, $\psi_{1, 4}$.}
%\label{fig:images_ordered}
\customlabel{fig:images_ordered}{5}
\end{figure}

@@

We have shown how symmetry and dimensionality reduction techniques can greatly accelerate the registration and temporal ordering of cross-sectional developmental imaging data.
%
Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
%
To our knowledge, temporal ordering of {\em Drosophila} images has not been automated to this extent.
%
Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.

Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
%
Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
%
Gene expression data from multiple patients has been studied in \cite{gupta2008extracting, qiu2011discovering};
here, each data point is a vector containing the expression levels of the various genes.
%
Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
%
These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
%
The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
expecting that these structures (the path of the travelling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate picture of the dynamical progression.
%
We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
%
Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
%
They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
%%%
%%%   we need help here
%%%
%
%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
%
%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.

The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
The main point to note here is that many registration algorithms require the locations of landmarks or features within each image.
%
For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
%
Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
%
In general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
%
Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
%
However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.

Angular synchronization and vector diffusion maps require no feature points and no template function.
%
These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\em same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
%
In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics. % (which underpins the development of vector diffusion map algorithms).
%
In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
%
Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
%
The main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.

Temporal ordering can also be accomplished without first registering the images, by constructing functions of the images which are invariant under translations and rotations. 
%
These functions are then used as the data in the diffusion maps calculation; since the functions do not change under translations and rotations, the resulting diffusion maps embedding is invariant to symmetries.
%
Such functions include the scale invariant feature transform \cite{lowe1999object, lowe2004distinctive}, the scattering transform \cite{mallat2012group}, and the bispectrum \cite{zhao2014rotationally}.
%
These methods have the disadvantage that we do not obtain a registered set of images, but only a temporal ordering.
%
However, such methods can often be faster for large data sets, since the functions are computed per image, rather than computing pairwise quantities in synchronization.

>>>>

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble the image features in Figure~\ref{fig:images_unordered}.
%
The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
%
We construct concentration profiles defined on a ring (as in Figure~\ref{subfig:one_image}), and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
{\bf We should discuss/mention that the eggs could be flipped.}
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
%
Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

@@

We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
%
Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
which make each pair have the maximum possible overlap (see {\it Materials and Methods}).
%%%
%%% so I do not like the figure with the vectors - it is correct, but people have to take the
%%% function, discretize and make it a vector in their heads, and may be "one step too much"
%%%
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
%
{\bf It is not clear how we are rotating the arrows (around the center vs. around the tail).}
%
From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
%
Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
The problem of finding these alignments is formulated as an optimization problem, and then relaxed to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
provides the rotation angles (up to a constant rotation) assigned to each data point (see {\it Materials and Methods}).
%
%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.
%
\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{(a) Set of vectors, each in a different orientation. The pariwise alignment angles are indicated. (b) The vectors from (a), each rotated so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in (a). (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges. (d) The data in (c), colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$. The color intensity is monotonic with the arclength, thus parameterizing the curve.  }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:synch1}{\ref{fig:schematics}a}
\customlabel{subfig:synch2}{\ref{fig:schematics}b}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}c}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}d}
\end{figure}
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
It is reassuring that the peaks appear aligned,
even though the algorithm uses no specific information about the existence or location of such peaks.
%
%We note once more that, since angular synchronization only uses pairwise rotation information,
%the optimal rotations are recovered up to a global rotation.

@@

After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
%
%Our main assumption is that most of the variation in the data correlates with time.
%
Our main assumptions are that the high-dimensional data (one dimension per discretization point in the concentration profiles/one dimension per color pixel intensity in the images) lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time.
%
Figure~\ref{subfig:dmaps_edges} shows a schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
%
We assume that the data points are ordered from youngest to oldest along the curve.
%
When our data are very high dimensional, we cannot extract this curve by eye.
%
We instead use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to automatically extract this one-dimensional curve.
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading (non-trivial) diffusion map coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect, and ordering the data by the leading diffusion map coordinate will effectively order data in time.

To uncover the diffusion map parameterization, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored based on the distance between the data points (using the diffusion kernel of the distance), so that pairs of points which are close have dark/strong edges connecting them.
%
Visually, the dark edges delineate the one-dimensional curve, whereas the light edges are less informative about the low-dimensional structure.
%
%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
%
Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
%
%This procedure has been named diffusion maps \cite{coifman2005geometric}.
%
%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we use diffusion maps to order them in time.
%
We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion map coordinate, $\phi_2$.
%
As expected, the intensity of the peak grows as a function of $\phi_2$ (since we constructed these data
to have a peak growing in time).
%
%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
%
%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.

@@

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the registrations {\em and} the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
%
The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion map coordinate.

@@

The same methods (with small modifications) can be applied directly to our two-dimensional images.
%
Instead of one-dimensional profiles (discretized to $100$-dimensional concentration vectors), we now work with
two color, two-dimensional images (discretized into $100 \times 100$ pixels/$20000$-dimensional vectors), and we register the images with respect to translations and rotations (the symmetry group is $ISO(2)$).
%
To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\em three-dimensional rotations} (symmetry group $SO(3)$; see {\it Materials and Methods}, {\it SI Appendix}).
%
We use vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
the results are shown in Figure~\ref{fig:images_ordered}.
%
Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of the leading vector diffusion map coordinate.

In these experiments, we are fortunate enough to be able to validate our  automatic registration and ordering using prior physical knowledge about the system.
%
For registration, it is known that the peak in the distribution of Dorsal (shown in green) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
%%%
%%% maybe you can also quantify this guy
%%%
%
%Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position.
%
Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
%
The two methods for registration and ordering can be visually compared through the sequences of one-dimensional profiles extracted from the circumference of the embryos (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
%
The general trends are consistent:
both green  and red peaks are aligned, and the red peaks grow as a function of time/leading vector diffusion map coordinate.
%
However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows an arguably better clustering of the ``outer" peaks in the top corners of the image.
%
%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
%
These peaks are known to arise later in development, and so it is reassuring that they appear at the end of the (recovered) trajectory.
%
Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
%
Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9158, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig5}
\caption{The images in Figure~\ref{subfig:images_unordered}, aligned and temporally ordered using vector diffusion maps. The images are sorted according to the value of the first non-trivial vector diffusion map component, $\psi_{1, 4}$.}
%\label{fig:images_ordered}
\customlabel{fig:images_ordered}{5}
\end{figure}

@@

We have shown how symmetry and dimensionality reduction techniques can successfully register and order cross-sectional developmental imaging data.
%
Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
%
To our knowledge, temporal ordering of {\em Drosophila} images has not been automated to this extent.
%
Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.

Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
%
Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
%
Gene expression data from multiple patients have been studied in \cite{gupta2008extracting, qiu2011discovering};
here, each data point is a vector containing the expression levels of the various genes.
%
Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
%
These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
%
The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
expecting that these structures (the path of the travelling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
%
We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
%
Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
%
They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
%
{\bf Yannis says ``careful'', he thinks this might be too strong of a statement.}
%%%
%%%   we need help here
%%%
%
%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
%
%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.

The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
The main point to note here is that many registration algorithms require definition of appropriate landmarks or features for the images and the location of these within each image.
%
For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
%
Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
%
In general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
%
Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
%
However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
%
{\bf Need to discuss consistency.}

Angular synchronization and vector diffusion maps require no feature points and no template function.
%
These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\em same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
%
In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics. % (which underpins the development of vector diffusion map algorithms).
%
In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
%
Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
%
Here, the main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.

Temporal ordering can also be accomplished without first registering the images, by constructing functions of the images which are invariant under translations and rotations. 
%
These functions are then used as the data in the diffusion maps calculation; since these representative functions do not change under translations and rotations, the resulting diffusion map embedding is invariant to symmetries.
%
Such functions include the scale invariant feature transform \cite{lowe1999object, lowe2004distinctive}, the scattering transform \cite{mallat2012group}, and the bispectrum \cite{zhao2014rotationally}.
%
These methods have the disadvantage that we do not obtain a registered set of images, but only a temporal ordering.
%
However, such methods can often be faster for large data sets, since the functions are computed per image, rather than computing pairwise quantities in synchronization.

In all of these data-driven methods, an essential question is how much and what type of data is needed to accurately reconstruct the dynamics.
%
For diffusion maps to temporally order the data, we require the developmental trajectory to be well-sampled, and that the principal direction of variability in the data (the coordinate which diffusion maps will recover) be monotonic in time.
%
In addition, the data should have spatial variability/structure for synchronization to successfully register the images.
%
In our example, the data set is sufficiently large so that the dynamics are well-sampled.
%
We have included both the red and the green channels so that the principal direction of variability is monotonic in time.
%
If we only used the green channel in our analysis, we could not successfully order the data using diffusion maps, as this signal does not change appreciably in time.
%
Both the red and green channels contain spatial variability which allows us to register the data points: the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
%
We include the green channel because the location of the red peaks is noisy, and so inclusion of the (less noisy) green channel allows us to more accurately register the images.
%
%For the methods we present here, we require a sufficient amount of data for the one-dimensional evolution curve to be ``seen'' by the algorithms (i.e., in Figure~\ref{subfig:dmaps_edges}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
%
%We also require the data to contain enough information so that the alignments and dynamics are apparent.
%
%For example, in the data presented in Figure~\ref{subfig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
%
%Therefore, these images contain information about rotational symmetries of the data as well as the dynamics of the data.
%
%If we removed the green signal, and only used the red signal in our analysis, the registration would not be as accurate.
%
%Furthermore, if we only used the green signal, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and the temporal ordering would be inaccurate.
%
We are confident that for many applications, these requirements - data which adequately samples the developmental trajectory, and data with sufficient spatial and temporal variability - are easily met.
%
In these cases, our methods allow for the rapid, automated analysis of imaging data sets to help uncover possibly complex dynamics and structure.
{\bf Something about smoothness in making movies}
%
%{\bf procrustes referece}
%{\bf allso the guy form Donoho who worried about how much to do syncjhronization}
